---
title: "Exercise 1"
subtitle: "for Data Analysis"
author: "Valentino Lazarevic - 1223211"
date: "12.03.2024"
output: pdf_document
---
```{r}
library(nycflights13)
data("flights")
attach(flights)
```
# Exercise 1
##  Show the variable air_time in a univariate scatterplot. What do you see?
## We can see that most of the flights have a duration time between 0 and 400 minutes. There are as well flights which have an air time of more than 570 minutes
```{r example_one, echo=FALSE, fig.align = 'center'}
stripchart(air_time, method = "jitter", xlab = "Air Time (minutes)", main = "Univariate scatterplot of air time", col = "blue")
```

# Exercise 2
##  Show the variable distance in a univariate scatterplot. What do you see?

## We can see that most flights have a distance  between 0 and 2750 miles, but there are as well some flights which have a lot more miles than usual flights.
```{r example_two, echo=FALSE, fig.align = 'center'}
stripchart(distance, method = "jitter", xlab = "Distance (miles)", main = "Univariate scatterplot of distance", col = "red")
```

# Exercise 3
##  Show air_time as well as distance in histograms, and adjust the histogram presentatation to get a good idea about the data structure. What do you see?

## We can see that the initial settings of both plots is a similar distribution of the data, but there is a dent in the curve of the distance plot
```{r example_three, echo=FALSE, fig.align = 'center'}
par(mfrow = c(1, 2))
hist(air_time, main = "Histogram of air_time", xlab = "Air Time (minutes)", breaks = "freedman-diaconis")
hist(distance, main = "Histogram of distance", xlab = "Distance (miles)", breaks = "freedman-diaconis")
```

# Exercise 4
##  Does the gap in the distribution of distance (shorter versus long distances) correspond to the gap in the distribution of air_time (shorter versus long)? Find an answer by using different color information in the plot(s).

## It does not correspond

```{r example_four, echo=FALSE, fig.align = 'center'}
plot(air_time, distance, col = ifelse(distance < 1000, "red", "blue"),
     main = "Air Time vs Distance", xlab = "Air Time (minutes)", ylab = "Distance (miles)")
legend("topright", legend = c("Short distance", "Long distance"),
       col = c("red", "blue"), pch = 1)
```

# Exercise 5
##  Investigate graphically if the distribution of arrival delay differs for short versus long flight distance. You could try to show two histograms in one plot using the same scale. A second histogram can be added to a plot by using the argument add=TRUE. What do you conclude?

## We conclude that short distance flights have arrival delay than long distance flights
```{r example_five, echo=FALSE, fig.align = 'center'}
hist(arr_delay[distance < 1000], main = "Arrival Delay for Short Distance Flights", xlab = "Arrival Delay (minutes)", xlim = c(-200, 300), col = "red", breaks = 20)
hist(arr_delay[distance > 1000], main = "Arrival Delay for Long Distance Flights", xlab = "Arrival Delay (minutes)", xlim = c(-200, 300), col = "blue", breaks = 20, add = TRUE)
legend("topright", legend = c("Short distance", "Long distance"), col = c("red", "blue"), pch = 15)
```

# Exercise 6
##   Do the same as in 5. based on the variable dep_delay.

## We conclude that short distance flights have arrival delay than long distance flights
```{r example_six, echo=FALSE, fig.align = 'center'}
hist(dep_delay[distance < 1000], main = "Departure Delay for Short Distance Flights", xlab = "Departure Delay (minutes)", xlim = c(-200, 300), col = "red", breaks = 20)
hist(dep_delay[distance > 1000], main = "Departure Delay for Long Distance Flights", xlab = "Departure Delay (minutes)", xlim = c(-200, 300), col = "blue", breaks = 20, add = TRUE)
legend("topright", legend = c("Short distance", "Long distance"), col = c("red", "blue"), pch = 15)
```

# Exercise 7
##   At which hour of the day can we expect the biggest arrival delays? When do we have the smallest arrival delays? How could you answer this question using simple plots?

## The hour with the biggest arrival delays is 21 and the hour with the smallest arrival delays is 7
```{r example_seven, echo=FALSE, fig.align = 'center'}
ad_median <- numeric(24)
ad_mean <- numeric(24)
for (i in 1:24) {
  flights_hour <- subset(flights, hour == i)
  if (length(flights_hour$hour) == 0) {
    ad_median[i] <- NA
    ad_mean[i] <- NA
  } else {
    ad_median[i] <- median(flights_hour$arr_delay, na.rm = TRUE)
    ad_mean[i] <- mean(flights_hour$arr_delay, na.rm = TRUE)
  }
}

plot(
  seq(1,24),
  ad_median,
  main = "Median Arrival Delays By Hours",
  xlab = "Hours",
  ylab = "Median Arrival Delays (minutes)",
)
axis(1, at = seq(1, 24, by = 1))

max_delay_hour <- which.max(ad_median)
points(max_delay_hour, ad_median[max_delay_hour], col = "red", pch = 16)
text(max_delay_hour, ad_median[max_delay_hour], max_delay_hour, pos = 1, col = "red")

min_delay_hour <- which.min(ad_median)
points(min_delay_hour, ad_median[min_delay_hour], col = "blue", pch = 16)
text(min_delay_hour, ad_median[min_delay_hour], min_delay_hour, pos = 3, col = "blue")
```

# Exercise 8
##  Is there is a difference of the distribution of arrival delays in the summer and winter months?

## There is a small difference between the summer and winter months like we can see in the histograms
```{r example_aight, echo=FALSE, fig.align = 'center'}
hist(arr_delay[month %in% c(6, 7, 8)], main = "Arrival Delay in Summer Months", xlab = "Arrival Delay (minutes)", col = "lightblue", breaks = 20)
hist(arr_delay[month %in% c(12, 1, 2)], main = "Arrival Delay in Winter Months", xlab = "Arrival Delay (minutes)", col = "lightgreen", breaks = 20, add = TRUE)
legend("topright", legend = c("Summer", "Winter"), fill = c("lightblue", "lightgreen"))
```

# Exercise 9
##  Install the R package epade and load the package. Use for the following task the function histogram.ade(), see also help file, to create superimposed histograms. Compare the airlines (carrier) AA (American Airlines) and WN (Southwest Airlines) concerning their distribution of arrival delays. You first need to create a data frame including the arrival delays of these carriers and a factor variable with the carriers. Use appropriate histogram parameters to make possible differences of the carriers clearly visible.
```{r example_nine, echo=FALSE, fig.align = 'center'}
library(epade)
carrier_data <- data.frame(
  carrier = factor(ifelse(carrier == "AA", "AA", "WN")),
  arr_delay = arr_delay
)

histogram.ade(
  data = carrier_data,
  x = "arr_delay",
  group = "carrier",
  main = "Arrival Delay Distribution by Carrier (AA vs. WN)",
  xlab = "Arrival Delay (minutes)",
  col = c("blue", "red"),
  breaks = "FD",  # Adjust the number of breaks for better visualization
  xlim = c(-100, 100),
  wall = 1
)
```

# Exercise 10
##   Compare visually the average arrival delays of the carriers AA and WN across the months. What do you conclude?

## We can see at the plot that American Airlines has on average more arrival delays than the Southwest Airlines
```{r example_ten, echo=FALSE, fig.align = 'center'}
aa <- subset(flights, carrier == "AA")
wn <- subset(flights, carrier == "WN")
aa_ad_mean <- numeric(12)
wn_ad_mean <- numeric(12)

for (i in 1:12) {
  aa_month <- subset(aa, aa$month == i)
  wn_month <- subset(wn, wn$month == i)
  aa_ad_mean[i] <- mean(aa_month$arr_delay, na.rm = TRUE)
  wn_ad_mean[i] <- mean(wn_month$arr_delay, na.rm = TRUE)
}

plot(
  seq(1, 12),
  aa_ad_mean,
  main = "Average Arrival Delays of Carriers Across the Months",
  col = "red",
  type = "l",
  xlab = "Months",
  ylab = "average Arrival Delays",
  ylim = c(-40, 40)
)
legend("topright", legend = c("AA", "WN"), col = c("blue", "red"), lty = 1)
axis(1, at = seq(1, 12, by = 1))
lines(seq(1,12), wn_ad_mean, col = "blue")
```
